Beyond May: Complexity-stability relationships \\ in high-dimensional dynamical systems

### 2023-04-23 papers/general-stability.tex

\documentclass[
 reprint,
%superscriptaddress,
%groupedaddress,
%unsortedaddress,
%runinaddress,
%frontmatterverbose, 
%preprint,
%preprintnumbers,
%nofootinbib,
%nobibnotes,
%bibnotes,
 amsmath,amssymb,
 aps,
%pra,
%prb,
%rmp,
%prstab,
%prstper,
%floatfix,
]{revtex4-2}

\usepackage{colortbl}
\usepackage{graphicx} 
\usepackage{booktabs}
\usepackage{caption}
\usepackage{notes2bib}
\newcommand{\red}{\textcolor{red}}

@@

Few mathematical arguments have influenced biological thinking like May's
prediction that large, complex systems cannot be stable~\cite{May1972}. 
It is, on the face of it, a perplexing conclusion. 
On the one hand, May's argument is so simple and compelling
that it hard not to conclude that it must hold universally,
and beyond biology~\cite{Haldane2011,Moran2019}. 
On the other, it is clear that at least some large, 
complex systems are stable---or else which stable patterns 
would ecology be studying in the first place? If anything, 
the opposite relationship between complexity and stability 
seems to hold empirically: rich, strongly coupled ecosystems 
(eg. rainforests) tend to be  stable over time, while sparser ones 
(eg. arctic environments) often exhibit large fluctuations, 
exctinctions, and invasions~\cite{Loreau2022,McCann2000,Hutchinson1959,Odum1959,MacArthur1955}.

@@

In ecology, many authors have sought to ease the tension between May's 
prediction and empirical observation by invoking effects not captured by 
dynamical systems with random 
coefficients~\cite{McCann2000,Chesson2000,Mougi2012,Rohr2014,Barabas2017,Grilli2017}. 
Here, we show that May's argument is itself incomplete: 
random dynamical systems do not necessarily imply that 
stability decreases with diversity or interaction 
strength---the opposite behavior is also possible, 
without the need for special or additional structure.

@@

Consider the dynamical system in $N$ variables
\begin{equation}\label{dynamics}
    \dot{x}_i = f(x_i) + \sum_{j}a_{ij}g(x_i)h(x_j) \, .
\end{equation}
Here $f(x_i)$ represents the self-dynamics of a population $i$, while $g(x_i)$ and $h(x_j)$ capture the interaction of $i$ with other populations. 
That interaction is weighted by a coefficient $a_{ij}$, such that $a_{ij} > 0$ (resp. $a_{ji} < 0$) implies a positive (resp. negative) effect of $j$ on the growth of $i$. 
In the following we refer to $f(x_i)$ as the ``production function'' and $g(x_i)$ as the ``response function''. 
This rather general setting has been used to study universality\cite{Barzel2013} and to construct 
minimal models for complex dynamics\cite{Barzel2015} ranging from biological~\cite{Alon2006,Karlebach2008,Skinner2012} 
to social~\cite{Pastor-Satorras2001,Hufnagel2004,Dodds2005} systems.

The classic generalized Lotka-Volterra (GLV) competition model corresponds to $f$, $g$, $h$ all linear. 
However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$. 
From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$. 
Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$~\cite{Brown2004}, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks. 
For reasons that are not currently understood, a similar pattern of growth appears to recur at the community level~\cite{Hatton2015,Hatton2023}. 
It has recently been shown appropriate to model predator-prey interactions with a square root laws ($g(x) \sim h(x) \sim x^{1/2}$)~\cite{Barbier2021,Mazzarisi2023}, for example.

@@

Consider now Eq.~\eqref{dynamics} with i.i.d. random 
interactions $a_{ij}$ with mean $\mu$ and standard 
deviation $\sigma$, with the diagonal elements $a_{ii}$ 
extracted from a distribution with mean $\mu_s$ 
possibly different from $\mu$ and same standard deviation
as for the off-diagonal elements.  
We can compute the Jacobian matrix at equilibrium
\begin{align}
    J_{ij}^* & = a_{ij}g_*(x_i)h_*'(x_j) \qquad \qquad \textrm{for} \ i\neq j \label{eq: jac off-diag}\\
    J_{ii}^* & = f_*'(x_i) - \frac{g_*'(x_i)f_*(x_i)}{g_*(x_i)} - a_{ii}g_*(x_i)h_*'(x_j) \ , \label{eq: jac diag}
\end{align}
where we used $\sum_{j}a_{ij}g_*(x_i)h_*(x_j)=-f_*(x_i)/g_*(x_i)$.
In order to investigates its spectral properties, 
we follow Stone~\cite{Stone2018} and use a
result in random matrix theory~\cite{Ahmadian2015}
which generalize the classical `circular law'~\cite{Potters2020}.

In Ref.~\cite{Ahmadian2015} Ahmadian \emph{et al.} 
consider matrices of the form $M + LSR$, where $M$,  
$L$ and $R$ are deterministic matrices, and $S$ 
is a random matrix with i.i.d. coefficients, 
zero mean and variance $\sigma^2$. They 
show that in the complex plane the boundary of the 
eigenspectrum of large matrices of this form is defined by
$\textrm{Tr}[(M_\zeta M_\zeta^\dagger)^{-1}]\geq \sigma^{-2}$ , 
where $M_\zeta = L^{-1}(\zeta I - M)R^{-1}$ and $\zeta\in\mathbb{C}$. 
If $L$, $R$ and $M$ are all diagonal $N\times N$ matrices, 
this condition simplifies to
\begin{equation}
    \sum_{i=1}^N\frac{(L_{i}R_{i})^2}{ \vert \zeta - M_{i}\vert^2 }\geq \sigma^{-2} \ .
\label{eq: domain}
\end{equation}

Now we decompose the interaction matrix as
$a = \mu \mathbf{1} + (\mu_s-\mu)I + S$,
with $\mathbf{1}$ the rank-one matrix with all entries equal to $1$
and $S$ a random matrix as defined above.
We can limit ourselves to consider 
$a = (\mu_s-\mu)I + S$ by relying on 
rank-one perturbation theory: we would only neglect outliers
uninfluential for the linear stbaility properties of the system (see Ref.~\cite{Stone2018}).
The Jacobian in Eqs.~\eqref{eq: jac off-diag} and \eqref{eq: jac diag} assumes therefore the form $M + LJR$ with
\begin{align}
    M & = \textrm{diag}\left(f_*'(\mathbf x) -
    \frac{g_*'(\mathbf x)f_*(\mathbf x)}{g_*(\mathbf x)}
    +(\mu_s-\mu)g_*(\mathbf x)h_*'(\mathbf x)\right) \ , \\
    L &= \textrm{diag}(g_*(\mathbf x)) \ , \qquad  
    R = \textrm{diag}(h_*'(\mathbf x)) \ . \nonumber
\end{align}
The domain of the eigenvalues of $J^*$, defined by
Eq.~\eqref{eq: domain}, 
first touches the right half-plane at $\zeta = 0$, 
hence linear stability requires   
\begin{equation}\label{eq: random-stability}
    \sum_{i=1}^N \cfrac{\left(\cfrac{g_*(x_i)h_*'(x_i)}{f_*(x_i)}\right)^2}{
        \Bigl |\left(\cfrac{f_*'(x_i)}{f_*(x_i)} -
        \cfrac{g_*'(x_i)}{g_*(x_i)}\right)
        +(\mu_s-\mu) \cfrac{g_*(x_i)h_*'(x_i)}{f_*(x_i)} \Bigl |^2}
    < \sigma^{-2}. 
\end{equation}
This expression can be considered the
generalization to random interaction of
the expression in Eq.~\eqref{homogeneous-general},
which appears again in the denominator of the summand.

@@

Following, e.g., Ref.~\cite{Roy2019} we can derive from 
the evolution equation
\begin{equation}
    \dot{x_i}=x_i^{\alpha} - x_i^{\beta}\sum_{j}a_{ij}x_j^{\gamma} \, ,
\label{eq: full abg}
\end{equation}
for large $N$, a dynamical mean field theory (DMFT) 
describing the entire ensemble of $N$ variables
thorugh the evolution of the representative random variable $x$
\begin{equation}
    \dot{x} = x^{\alpha}-a_sx^{\beta+\gamma}-x^{\beta}\left( \mu N \langle x^{\gamma}\rangle + \sigma \sqrt{N} \eta\right) \, ,
\label{eq: dmft}
\end{equation}
where $a_s$ is a random variable with the statistics of $a_{ii}$,
$\langle . \rangle$ stands for expectation value
and $\eta$ is a gaussian variable
with zero mean and correlation 
$\langle x^{\gamma}(t)x^{\gamma}(s)\rangle$. 
The averages and correlations have to be computed self-consistently.
If the system admits an equilibrium, for 
$t\to\infty$ each of the initial $N$ degrees 
of freedom will reach a final finite value. 
The effective random variable describing their statistics
also become time-independent,
$\lim_{t\to\infty}x(t)=x_*$ and we can write
\begin{equation}
    x_*^{\alpha-\beta} - a_s x_*^{\gamma}= \left( \mu N \langle x_*^{\gamma}\rangle + \sigma \sqrt{N\langle x_*^{2\gamma}\rangle}\xi\right) \, ,
\end{equation} 
where $\xi$ is a standard normal random variable (i.e. $\xi\sim\mathcal{N}(0,1)$).
The equation above can be solved for $x_*$ for specific values of
$(\alpha-\beta)/\gamma$, or, in general, if we set $a_s=0$, which amounts to include
all the self-interaction in the nonlinear produciton term
$x_*^{\alpha}$. In the following, for cases which
falls in the "anti-May" phase, we show that numerical results
for non-vanishing diagonal part of the
interaction matrix and equal to the
off-diagonal part are well approximated by the results obtained
setting $a_s=0$.

@@

Let us focus on the specific choice 
$\alpha=1$, $\beta=1.5$ and $\gamma=1$, with
$a_s=0$ in order to leverage the equilibrium solution in Eq.~\eqref{eq: dist general}.
In this case the first and second moments, 
to be self-consistently computed, formally diverges. This is not truly porblematic in practice
because we are dealing with large, but \emph{finite} populations.
Following~\cite{Cui2020,Hatton2023} and assuming
$\sigma \sqrt{N\langle x_*^2\rangle}\ll \mu N \langle x_* \rangle$,
we can Taylor expand to the first-order in 
$\sigma \sqrt{N\langle x_*^2\rangle}$ the cavity 
solution~\eqref{eq: cavity solution} and 
approximate to Eq.~\eqref{eq: dist general} with
a gaussian distribuiton
with moments $\langle x_*\rangle=(\mu N)^{-2/3}$ and
$\langle x_*^2\rangle=(\mu N)^{-4/3}/(1-4(\mu N)^{-2}\sigma^2N)$.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=.45\textwidth]{figs/cavity.pdf}
    \caption{The equilibrium distribution for a realization
    of the system described by Eq.~\eqref{eq: full abg} 
    is accurately reproduced by the cavity solution in 
    Eq.\eqref{eq: dist general} and its gaussian approximation.
    For the simulation we used  $\alpha=1$, $\beta=1.5$,
    $\gamma=1$, $N=50$, $\mu=\mu_s=10^{-2}$, $\sigma=10^{-2}/2$.}
    \label{fig: cavity sol.}
\end{figure}

Equipped with the equilibirum distribution, we can compute
the critical value of heterogeneity 
$\sigma_c$, for a given $N$ and $\mu$,
defined by taking the equality in 
Eq.~\eqref{eq: random-stability int}. 
For $\sigma$ above this value the system becomes
unstable. In Fig.~\ref{fig: stability line + sims}
the result is portrayed in the $(\sigma \sqrt{N},\mu N)$ plane,
compared with results from simulations where stability is defined
as stable, full, coexistence of the system.
Everything else left fixed, an increase in the number of
degrees of freedom $N$ will move the system along a square-root
trajectory, bringing it into the stable regime or keeping it
in it if already stable. In other words, we are in the
"anti-May" phase. The stability boundary for GLV, by contrast, is touched
at $\sigma \sqrt{N} = \mu N/N + \mu_s$, which, for $N\to \infty$
gives an horizontal line in the $(\sigma \sqrt{N},\mu N)$ plane,
defined by $\sigma_c\sqrt{N} = \mu_s$.

\begin{figure}[h!]
    \centering
    \includegraphics[width=.45\textwidth]{figs/beta1_5-S50-N10-diversity.pdf}
    \caption{For a system in the "anti-May" phase an increase in $N$,
    corresponding to moving along $\sqrt{N}$ in the $(\sigma \sqrt{N},\mu N)$ plane,
    will be stabilizing. Here we show results from simulation of system~\eqref{eq: full abg},
    where stability is defined as full stable coexistence, overleyed with the critical line
    obtained through the cavity method starting from Eq.\eqref{eq: random-stability int} 
    (see main text). The values of the parameters are $\alpha=1$, $\beta=1.5$,
    $\gamma=1$, $N=50$ and $10$ replicates for the simulations.}
    \label{fig: stability line + sims}
\end{figure}

Figure~\ref{fig: alpha-beta} portrays results of simulaitons 
for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, 
at fixed $\mu N$ and $\gamma = 1$ (further details in caption), 
defined as the value above which
full stable coexistence (i.e. 'stability'
as defined for Fig.\ref{fig: stability line + sims}) 
is less probable than $50\%$.
In these simulations we are using $\mu_s = \mu$, which result
in unstable outcomes for $\beta<\alpha$ while
a certain degree of heterogeneity in the interactions
admits admits stability for $\beta>\alpha$, the higher
the larger is the difference between $\beta$ and $\alpha$.
Notice that the existence of a $\sigma_c\sqrt{N}$ above which
instability is triggered does not imply that increasing $N$ will
destabilize the system, because also $\mu N$ would change.
Rather, for all the value of $\beta$ and $\alpha$ in the
upper triangle, the picture is the same as the one depicted in 
Fig.~\ref{fig: stability line + sims}: we are in "anti-May"
phase.

\begin{figure}[h!]
    \centering
    \includegraphics[width=.45\textwidth]{figs/alpha-beta.pdf}
    \caption{The "May" and "anti-May" phases described for the
    case of uniform interactions are robust for random interactions.
    Here we simulated system~\eqref{eq: full abg} for a fixed value of
    $\mu=\mu_s=10^{-2}$, $\gamma=1$, $N=100$ and $100$ replicates 
    and evaluated $\sigma_c$, defined here as
    the value above which full stable coexistence is less probable than 50\%,
    at varying $\alpha$ and $\beta$. No amount of heterogeneity is allowed in the
    lower triangle because we have $\mu=\mu_s$, while an increasing value of $\sigma_c$
    is found as $\beta$ becomes bigger than $\alpha$.
    We stress that the existence of a finite $\sigma_c\sqrt{N}$ above which
    instability is triggered also in the upper triangle,
    does not imply that increasing $N$ will
    destabilize the system, because also $\mu N$ would change (see main text).
    }
    \label{fig: alpha-beta}
\end{figure}

@@

\clearpage

\bibliography{general-stability-bib}

\bibliographystyle{unsrt}

>>>>

\documentclass[%
 reprint,
%superscriptaddress,
%groupedaddress,
%unsortedaddress,
%runinaddress,
%frontmatterverbose, 
%preprint,
%preprintnumbers,
%nofootinbib,
%nobibnotes,
%bibnotes,
 amsmath,amssymb,
 aps,
%pra,
%prb,
%rmp,
%prstab,
%prstper,
%floatfix,
]{revtex4-2}

\usepackage{colortbl}
\usepackage{graphicx} 
\usepackage{booktabs}
\usepackage{caption}
\newcommand{\red}{\textcolor{red}}

@@

Few mathematical arguments have influenced biological thinking like May's prediction that large, complex systems cannot be stable. It is, on the face of it, a perplexing conclusion. On the one hand, May's argument is so simple and compelling that it hard not to conclude that it must hold universally. On the other, it is clear that at least some large, complex systems are stable---or else which stable patterns would biology be studying in the first place? If anything, the opposite relationship between complexity and stability seems to hold empirically: rich, strongly coupled ecosystems (eg. rainforests) tend to be  stable over time, while sparser ones (eg. arctic environments) often exhibit large fluctuations, exctinctions, and invasions.

@@

In ecology, many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients, such as spatial or food-web structure, time delays, etc. 
Here, we show that May's argument is itself incomplete: random dynamical systems do not necessarily imply that stability decreases with diversity or interaction strength---the opposite behavior is also possible, without the need for special or additional structure.

@@

Consider the dynamical system in $N$ variables
\begin{equation}\label{dynamics}
    \dot{x}_i = f(x_i) + \sum_{j}a_{ij}g(x_i)h(x_j) \, .
\end{equation}
Here $f(x_i)$ represents the contribution of a population $i$ to its own growth, while $g(x_i)$ and $h(x_j)$ capture the interaction of $i$ with other populations. 
That interaction is weighted by a coefficient $a_{ij}$, such that $a_{ij} > 0$ (resp. $a_{ji} < 0$) implies a positive (resp. negative) effect of $j$ on the growth of $i$. 
In the following we refer to $f(x_i)$ as the ``production function'' and $g(x_i)$ as the ``response function''.

The classic generalized Lotka-Volterra (GLV) competition model corresponds to $f$, $g$, $h$ all linear. 
However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$. 
From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$. 
Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks. 
For reasons that are not currently understood, a similar pattern of growth appears to recur at the community level. 
It is also common to model predator-prey interactions with a square root laws ($g(x) \sim h(x) \sim x^{1/2}$), for example.

@@

Consider now Eq.~\eqref{dynamics} with i.i.d. random 
interactions $a_{ij}$ with mean $\mu$ and standard 
deviation $\sigma$, with the diagonal elements $a_{ii}$ 
extracted from a distribution with mean $\mu_s$ 
possibly different from $\mu$ and same standard deviation
as for the off-diagonal elements.  
We can compute the Jacobian matrix at equilibrium
\begin{align}
    J_{ij}^* & = a_{ij}g_*(x_i)h_*'(x_j) \qquad \qquad \textrm{for} \ i\neq j \label{eq: jac off-diag}\\
    J_{ii}^* & = f_*'(x_i) - \frac{g_*'(x_i)f_*(x_i)}{g_*(x_i)} - a_{ii}g_*(x_i)h_*'(x_j) \ , \label{eq: jac diag}
\end{align}
where we used $\sum_{j}a_{ij}g_*(x_i)h_*(x_j)=-f_*(x_i)/g_*(x_i)$.
In order to investigates its spectral properties, 
we follow Stone~[Stone2018] and use a
result in random matrix theory~[Ahmadian2015]
which generalize the classical `circular law'~[Circular].

In Ref.~[Ahmadian2015] Ahmadian \emph{et al.} 
consider matrices of the form $M + LSR$, where $M$,  
$L$ and $R$ are deterministic matrices, and $S$ 
is a random matrix with i.i.d. coefficients, 
zero mean and variance $\sigma^2$. They 
show that in the complex plane the boundary of the 
eigenspectrum of large matrices of this form is defined by
$\textrm{Tr}[(M_\zeta M_\zeta^\dagger)^{-1}]\geq \sigma^{-2}$ , 
where $M_\zeta = L^{-1}(\zeta I - M)R^{-1}$ and $\zeta\in\mathbb{C}$. 
If $L$, $R$ and $M$ are all diagonal $N\times N$ matrices, 
this condition simplifies to
\begin{equation}
    \sum_{i=1}^N\frac{(L_{i}R_{i})^2}{ \vert \zeta - M_{i}\vert^2 }\geq \sigma^{-2} \ .
\label{eq: domain}
\end{equation}

Now we decompose the interaction matrix as
$a = \mu \mathbf{1} + (\mu_s-\mu)I + S$,
with $\mathbf{1}$ the rank-one matrix with all entries equal to $1$
and $S$ a random matrix as defined above.
We can limit ourselves to consider 
$a = (\mu_s-\mu)I + S$ by relying on 
rank-one perturbation theory: we would only neglect outliers
uninfluential for the linear stbaility properties of the system (see Ref.~[Stone2018]).
The Jacobian in Eqs.~\eqref{eq: jac off-diag} and \eqref{eq: jac diag} assumes therefore the form $M + LJR$ with
\begin{align}
    M & = \textrm{diag}\left(f_*'(\mathbf x) -
    \frac{g_*'(\mathbf x)f_*(\mathbf x)}{g_*(\mathbf x)}
    +(\mu_s-\mu)g_*(\mathbf x)h_*'(\mathbf x)\right) \ , \\
    L &= \textrm{diag}(g_*(\mathbf x)) \ , \qquad  
    R = \textrm{diag}(h_*'(\mathbf x)) \ . \nonumber
\end{align}
The domain of the eigenvalues of $J^*$, defined by
Eq.~\eqref{eq: domain}, 
first touches the right half-plane at $\zeta = 0$, 
hence linear stability requires   
\begin{equation}\label{eq: random-stability}
    \sum_{i=1}^N \cfrac{\left(\cfrac{g_*(x_i)h_*'(x_i)}{f_*(x_i)}\right)^2}{
        \Bigl |\left(\cfrac{f_*'(x_i)}{f_*(x_i)} -
        \cfrac{g_*'(x_i)}{g_*(x_i)}\right)
        +(\mu_s-\mu) \cfrac{g_*(x_i)h_*'(x_i)}{f_*(x_i)} \Bigl |^2}
    < \sigma^{-2}. 
\end{equation}
This expression can be considered the
generalization to random interaction of
the expression in Eq.~\eqref{homogeneous-general},
which appears again in the denominator of the summand.

@@

Following, e.g., Ref.~[Roy2019] we can derive from 
the evolution equation
\begin{equation}
    \dot{x_i}=x_i^{\alpha} - x_i^{\beta}\sum_{j}a_{ij}x_j^{\gamma} \, ,
\label{eq: full abg}
\end{equation}
for large $N$, a dynamical mean field theory (DMFT) 
describing the entire ensemble of $N$ variables
thorugh the evolution of the representative random variable $x$
\begin{equation}
    \dot{x} = x^{\alpha}-a_sx^{\beta+\gamma}-x^{\beta}\left( \mu N \langle x^{\gamma}\rangle + \sigma \sqrt{N} \eta\right) \, ,
\label{eq: dmft}
\end{equation}
where $a_s$ is a random variable with the statistics of $a_{ii}$,
$\langle . \rangle$ stands for expectation value
and $\eta$ is a gaussian variable
with zero mean and correlation 
$\langle x^{\gamma}(t)x^{\gamma}(s)\rangle$. 
The averages and correlations have to be computed self-consistently.
If the system admits an equilibrium, for 
$t\to\infty$ each of the initial $N$ degrees 
of freedom will reach a final finite value. 
The effective random variable describing their statistics
also become time-independent,
$\lim_{t\to\infty}x(t)=x_*$ and we can write
\begin{equation}
    x_*^{\alpha-\beta} - a_s x_*^{\gamma}= \left( \mu N \langle x_*^{\gamma}\rangle + \sigma \sqrt{N\langle x_*^{2\gamma}\rangle}\xi\right) \, ,
\end{equation} 
where $\xi$ is a standard normal random variable (i.e. $\xi\sim\mathcal{N}(0,1)$).
The equation above can be solved for $x_*$ for specific values of
$(\alpha-\beta)/\gamma$, or, in general, if we set $a_s=0$, which amounts to include
all the self-interaction in the nonlinear produciton term
$x_*^{\alpha}$. In the following, for cases which
falls in the 'anti-May' phase, we show that numerical results
for non-vanishing diagonal part of the
interaction matrix and equal to the
off-diagonal part are well approximated by the results obtained
setting $a_s=0$.

@@

Let us focus on the specific choice 
$\alpha=1$, $\beta=1.5$ and $\gamma=1$, with
$a_s=0$ in order to leverage the equilibrium solution in Eq.~\eqref{eq: dist general}.
In this case the first and second moments, 
to be self-consistently computed, formally diverges. This is not truly porblematic in practice
because we are dealing with large, but \emph{finite} populations.
Following~[Cui2020,Us2023] and assuming
$\sigma \sqrt{N\langle x_*^2\rangle}\ll \mu N \langle x_* \rangle$,
we can Taylor expand to the first-order in 
$\sigma \sqrt{N\langle x_*^2\rangle}$ the cavity 
solution~\eqref{eq: cavity solution} and 
approximate to Eq.~\eqref{eq: dist general} with
a gaussian distribuiton
with moments $\langle x_*\rangle=(\mu N)^{-2/3}$ and
$\langle x_*^2\rangle=(\mu N)^{-4/3}/(1-4(\mu N)^{-2}\sigma^2N)$.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=.45\textwidth]{figs/cavity.pdf}
    \label{fig: cavity sol.}
    \caption{The equilibrium distribution for a realization
    of the system described by Eq.~\eqref{eq: full abg} 
    is accurately reproduced by the cavity solution in 
    Eq.\eqref{eq: dist general} and its gaussian approximation.
    For the simulation we used  $\alpha=1$, $\beta=1.5$,
    $\gamma=1$, $N=50$, $\mu=\mu_s=10^{-2}$, $\sigma=10^{-2}/2$.}
\end{figure}

Equipped with the equilibirum distribution, we can compute
the critical value of heterogeneity 
$\sigma_c$, for a given $N$ and $\mu$,
defined by taking the equality in 
Eq.~\eqref{eq: random-stability int}. 
For $\sigma$ above this value the system becomes
unstable. In Fig.~\ref{fig: stability line + sims}
the result is portrayed in the $(\sigma \sqrt{N},\mu N)$ plane,
compared with results from simulations where stability is defined
as stable, full coexistence of the system.
Everything else left fixed, an increase in the number of
degrees of freedom $N$ will move the system along a square-root
trajectory, bringing it into the stable regime or keeping it
in it if already stable. In other words, we are in the
'anti-May' phase. The stability boundary for GLV, by contrast, is touched
at $\sigma \sqrt{N} = \mu N/N + \mu_s$, which, for $N\to \infty$
gives an horizontal line in the $(\sigma \sqrt{N},\mu N)$ plane,
defined by $\sigma_c\sqrt{N} = \mu_s$.

\begin{figure}[h!]
    \centering
    \includegraphics[width=.45\textwidth]{figs/beta1_5-S50-N10-diversity.pdf}
    \caption{... $\alpha=1$, $\beta=1.5$,
    $\gamma=1$, $N=50$ and $10$ replicates.}
    \label{fig: stability line + sims}
\end{figure}

Figure~\ref{fig: alpha-beta} portrays results of simulaitons 
for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, 
at fixed $\mu N$ and $\gamma = 1$ (further details in caption), 
defined as the value above which
full stable coexistence (i.e. 'stability'
as defined for Fig.\ref{fig: stability line + sims}) 
is less probable then $50\%$.
In these simulations we are using $\mu_s = \mu$, which result
in unstable outcomes for $\beta<\alpha$ while
a certain degree of heterogeneity in the interactions
admits admits stability for $\beta>\alpha$, the higher
the larger is the difference between $\beta$ and $\alpha$.
Notice that the existence of a $\sigma_c\sqrt{N}$ above which
instability is triggered does not imply that increasing $N$ will
destabilize the system, because also $\mu N$ would change.
Rather, for all the value of $\beta$ and $\alpha$ in the
upper triangle, the picture is the same as the one depicted in 
Fig.~\ref{fig: stability line + sims}: we are in 'anti-May'
phase.

\begin{figure}[h!]
    \centering
    \includegraphics[width=.45\textwidth]{figs/alpha-beta.pdf}
    \caption{...
    $\mu=\mu_s=10^{-2}$, $N=100$ and $100$ replicates.}
    \label{fig: alpha-beta}
\end{figure}

@@
